<<<<<<< HEAD
---
title: "Final Project"
author: "Emily Mynar, Jack McCormick, & Pierre Beaurang"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.width=8, 
                      fig.height=5)


# Load packages
pacman::p_load(dplyr, ggplot2, tidyverse, ggthemes, maps, ggrepel)

# Set default theme
theme_set(theme_bw())

# Read in files
parks <- read.csv('parks.csv') |>
  janitor::clean_names()

species <- read.csv('species.csv') |>
  janitor::clean_names() 

tibble(species)
tibble(parks)
```
### Data Cleaning

Our data comes from two data sets:

1) **parks:** contains general information (size, latitude, longitude, etc.) about each National Park in the United States. This will be largely supplemental, but provides important contextual information we will use later

2) **species:** contains a record of each species recorded in a certain National Park. This will be our main data set.

Along with the true taxonomic groupings of order and family, the data also includes a *'category'* variable, which is not equivalent to any true taxonomic grouping, but instead consists of informal groupings which are recognizable to the layperson. Let's take a look at them now, since we'll be using these groups for much of our analysis.
```{r categories}

# Display species categories
categories <-
  species |>
  group_by(category) |>
  summarize(category_count = n()) |>
  select(category)

tibble(categories)
```
14 categories, but for our purposes, we only care about the **animal** diversity of these parks. As such, we will go ahead and remove *algae*, *fungi*, *Nonvascular plants*, and *Vascular plants*.  

As we began analyzing our data, we realized that reporting of insects was inconsistent, and skewed our data heavily in the wrong direction. As such, we decided to limit our scope to **vertebrate** biodiversity rather than animal diversity. To accomplish this, we removed the *insect* group which was causing problems, as well as all other invertebrate categories *Invertebrate*, *Crab/Lobster/Shrimp*, *Slug/Snail* and *Spider/Scorpion*

We'll also implement a handful of other changes to our data: removing record status and seasonality, which are not useful for our driving question, adding a genus category, which is taken from the scientific name, and shortening the names of parks to remove the suffix for clarity.

```{r data cleaning}

species_clean <-
  species |>
  # Remove record status and seasonality
  select(-record_status, -seasonality, -x) |>
  # Filter non-animal species,
  # Filter invertebrate species
  filter(!category %in% c('Algae', 'Fungi', 'Nonvascular Plant', 'Vascular Plant', 'Insect', 'Invertebrate', 'Crab/Lobster/Shrimp', 'Slug/Snail', 'Spider/Scorpion')) |>
  # Add genus column
  mutate(genus = str_split(scientific_name, " ", simplify = T)[ , 1]) |>
  # Reorder columns
  relocate(genus, .before = scientific_name) |>
  # Make park names shorter
  mutate(park_name = str_remove(park_name, 'National Parks'),
         park_name = str_remove(park_name, 'National Park'),
         park_name = str_remove(park_name, 'and Preserve'))

tibble(species_clean)


parks_clean <-
  parks |>
  # Make park names shorter
  mutate(park_name = str_remove(park_name, 'National Parks'),
         park_name = str_remove(park_name, 'National Park'),
         park_name = str_remove(park_name, 'and Preserve'))

tibble(parks_clean)
```
Lastly, we'll combine the two data sets and remove the park code.

We have decided to restrict our data to the contiguous United States, for a number of reasons. Alaska and Hawaii have populations that are so different from the rest of the country that they are hardly comparable. Removing them makes our data far less complex, as it removes species that are found nowhere else in the country. It also serves the secondary purpose of making the data mapping easier. 

```{r combined dataset}
species_park <-
  species_clean |>
  # Add park info to species dataset
  left_join(y = parks_clean, by = 'park_name') |>
  # Remove park code
  select(-park_code) |>
  # Remove parks in Alaska or Hawaii for mapping purposes
  filter(!state %in% c('AK', 'HI')) 
  
tibble(species_park)
```

Here are our updated species, categories, which will be used throughout the rest of the analysis. 

```{r updated categories}

# Display updated species categories
categories_clean <-
  species_clean |>
  group_by(category) |>
  summarize(category_count = n()) |>
  select(category)

tibble(categories_clean)

```
### Data Analysis

Because our categories are not true taxonomic groupings, we need some context to the diversity within each. Lets examine orders first.

```{r overall taxanomic categories proportion}

category_prop <-
  
  species_clean |>
  group_by(category) |>
  summarize(number_cat = n()) |>
  mutate(number_tot = sum(number_cat),
         prop = number_cat/number_tot) |>
  select(category, prop)

category_prop |>
  
  ggplot(mapping = aes(
    y = prop,
    x = "",
    fill = category
  )) +

  geom_col(position = "fill",
           color = "black") +
  
  labs(x = NULL,
       y = "Proportion by Category",
       fill = "Category",
       title = "Proportion of Each Taxanomic Category")

#going to be working on this below, will probably have to do some small multiples shenanigans bc there are way too many orders
# TODO: we honestly may not need this, or at least we can wait until we get feedback from him on if we need more initial graphs/info
  species_clean |>
  group_by(category, order) |>
  summarize(number_ord = n()) |>
  mutate(number_tot = sum(number_ord),
         prop = number_ord/number_tot) |>
  select(category, order, prop) |>
  
  ggplot(mapping = aes(
    x = prop,
    y = "",
    fill = order
  )) +
    
    geom_col(position = "fill",
           color = "black") +
    labs(y = NULL,
       x = "Proportion by Order",
       fill = "Category",
       title = "Proportion of Each Order") +
    facet_wrap(facets = '~category') +
    
    # TODO: Maybe we should do number of different orders within each category rather than proportion
    # I'm trying that below
    theme(legend.position = 'none')


```
As seen, the vast majority of species observed fall into the **bird** categories. This can be explained by a few different hypotheses:

1) Because we removed seasonality, we are looking at species spotted in the park at any time of year, including species that are strictly migratory. Due to their migratory nature, the average bird species is often has a wider range of observation than that of an equivalent reptile, amphibian or mammal. 

2) Additionally, due to a variety of factors, birds are often have more comprehensive records than any other group. The MAPS program, created by the Institute for Bird Populations has gone a long way to creating distribution patterns across the United States, and the role of citizen science, on programs like eBird or iNaturalist create lists of observed species without the need for a monitoring program run by the park. 

3) Birds naturally have a high diversity, and occur across ecosystems. As such, every park is likely to have birds present, while fish, amphibians, reptiles, and mammals are far more restricted to certain habitats. 
```{r species by park}

ggplot(data = species_park |> group_by(park_name),
       mapping = aes(x = fct_infreq(park_name))) +
  
  geom_bar(fill = 'deepskyblue3') +
  
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1,
                                   size = 7),
        plot.title = element_text(size = 12,
                                  hjust = 0.5,
                                  face = 'bold'),
        axis.title = element_text(size = 10)) +
  
  labs(x = 'Park name',
       y = 'Number of species',
       fill = 'Category',
       title = 'Number of species in each park') +
  
  scale_y_continuous(expand = c(0,0, 0.05, 0)) 


````


```{r species richness order}

species_per_park <-
  species_park |>
  group_by(park_name, latitude, longitude, state, acres) |>
  summarise(num_species = n()) |>
  arrange(desc(num_species))

tibble(species_per_park)

park_order <-
  species_per_park$park_name


```


```{r taxonomic categories by park}

categories_by_park <-
  species_park |>
  group_by(park_name, category) |>
  summarise(cat_sum = n())

ggplot(data = categories_by_park,
       mapping = aes(x = factor(park_name,
                                levels = park_order))) +
  
  geom_col(mapping = aes(y = cat_sum,
                         fill = category),
           position = 'stack') +
  
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1,
                                   size = 7),
        plot.title = element_text(size = 12,
                                  hjust = 0.5,
                                  face = 'bold'),
        axis.title = element_text(size = 10)) +
  
  labs(x = 'Park name',
       y = 'Number of species',
       fill = 'Category',
       title = 'Number of species in each park by category') +
  
  scale_y_continuous(expand = c(0,0, 0.05, 0)) 



# Same plot but by proportion
ggplot(data = categories_by_park,
       mapping = aes(x = factor(park_name,
                                 levels = park_order),
                     y = cat_sum,
                     fill = category)) +
  
  geom_col(position = 'fill') +
  
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1,
                                   size = 7),
        plot.title = element_text(size = 12,
                                  hjust = 0.5,
                                  face = 'bold'),
        axis.title = element_text(size = 10)) +
  
  labs(x = 'Park name',
       y = 'Proportion of species',
       fill = 'Category',
       title = 'Proportion of species in each park by category') +
  
  scale_y_continuous(labels = scales::percent,
                     expand = c(0,0)) 
  
  

  

```
```{r map outline}
map_outline <-
  ggplot(data = map_data(map = 'state'),
         mapping = aes(x = long,
                       y = lat,
                       group = group)) +
  
  geom_polygon(fill = 'white',
               color = 'grey30') +
  
  theme_map() +
  
  coord_map(projection = "albers", 
            lat0 = 39, lat1 = 45) +
  
  scale_x_continuous(expand = c(0,0)) +
  
  scale_y_continuous(expand = c(0,0))

```


```{r species richness map}
 richness_map <-
  
  map_outline +
  
  geom_point(data = species_per_park,
           mapping = aes(x = longitude,
                        y = latitude,
                       group = park_name,
                       size = num_species,
                       color = state)) +
  
  geom_text_repel(data = species_per_park,
            mapping = aes(x = longitude,
                          y = latitude,
                          group = park_name,
                          label = park_name),
            hjust = 0.1,
            size = 1.8,
            fontface= 'bold',
            max.overlaps = 15) +
  
  scale_size(breaks = c(500, 1000, 1500, 2000, 2500, 3000)) +
  
  labs(title = 'Species richness by park',
       size = 'Number of species') +
  
  guides(color = F) +
  
  theme(legend.position = 'bottom',
        legend.direction = 'horizontal',
        legend.box.background = element_rect(color = 'black'),
        plot.title = element_text(size = 12,
                                  hjust = 0.5,
                                  face = 'bold'))
  
  
richness_map
```


```{r park size map}
richness_map <-
  
  map_outline +
  
  geom_point(data = species_per_park,
           mapping = aes(x = longitude,
                        y = latitude,
                       group = park_name,
                       size = acres,
                       color = state)) +
  
  geom_text_repel(data = species_per_park,
            mapping = aes(x = longitude,
                          y = latitude,
                          group = park_name,
                          label = park_name),
            hjust = 0.1,
            size = 1.8,
            fontface= 'bold',
            max.overlaps = 15) +
  
  # TODO: Should figure out better breaks maybe?
  scale_size(breaks = seq(from = 100000, to = 2500000, by = 200000)) +
  
  labs(title = 'Park size',
       size = 'Acreage') +
  
  guides(color = F) +
  
  theme(legend.position = 'bottom',
        legend.direction = 'horizontal',
        legend.box.background = element_rect(color = 'black'),
        plot.title = element_text(size = 12,
                                  hjust = 0.5,
                                  face = 'bold'))
  
  
richness_map
```

```{r scatter plot of size vs richness}
# very much in progress duh
ggplot(data = species_per_park,
       mapping = aes(x = acres,
                     y = num_species)) +
  
  geom_point()
```

```{r size richness correlation}
species_per_park |>
  mutate(
    number = "num_species"
  ) -> 
  species_corr 
  


ggcorr(
  data = species_corr,
  low = "red3",
  mid = "white",
  high = "blue3",
  label = T,
  label_round = 2
  
)
```

=======
---
title: "Final Project"
author: "Emily Mynar, Jack McCormick, & Pierre Beaurang"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.width=8, 
                      fig.height=5)


# Load packages
pacman::p_load(dplyr, ggplot2, tidyverse, ggthemes, maps, ggrepel)

# Set default theme
theme_set(theme_bw())

# Read in files
parks <- read.csv('parks.csv') |>
  janitor::clean_names()

species <- read.csv('species.csv') |>
  janitor::clean_names() 

tibble(species)
tibble(parks)
```
### Data Cleaning

Our data comes from two data sets:

1) **parks:** contains general information (size, latitude, longitude, etc.) about each National Park in the United States. This will be largely supplemental, but provides important contextual information we will use later

2) **species:** contains a record of each species recorded in a certain National Park. This will be our main data set.

Along with the true taxonomic groupings of order and family, the data also includes a *'category'* variable, which is not equivalent to any true taxonomic grouping, but instead consists of informal groupings which are recognizable to the layperson. Let's take a look at them now, since we'll be using these groups for much of our analysis.
```{r categories}

# Display species categories
categories <-
  species |>
  group_by(category) |>
  summarize(category_count = n()) |>
  select(category)

tibble(categories)
```
14 categories, but for our purposes, we only care about the **animal** diversity of these parks. As such, we will go ahead and remove *algae*, *fungi*, *Nonvascular plants*, and *Vascular plants*.  

As we began analyzing our data, we realized that reporting of insects was inconsistent, and skewed our data heavily in the wrong direction. As such, we decided to limit our scope to **vertebrate** biodiversity rather than animal diversity. To accomplish this, we removed the *insect* group which was causing problems, as well as all other invertebrate categories *Invertebrate*, *Crab/Lobster/Shrimp*, *Slug/Snail* and *Spider/Scorpion*

We'll also implement a handful of other changes to our data: removing record status and seasonality, which are not useful for our driving question, adding a genus category, which is taken from the scientific name, and shortening the names of parks to remove the suffix for clarity.

```{r data cleaning}

species_clean <-
  species |>
  # Remove record status and seasonality
  select(-record_status, -seasonality, -x) |>
  # Filter non-animal species,
  # Filter invertebrate species
  filter(!category %in% c('Algae', 'Fungi', 'Nonvascular Plant', 'Vascular Plant', 'Insect', 'Invertebrate', 'Crab/Lobster/Shrimp', 'Slug/Snail', 'Spider/Scorpion')) |>
  # Add genus column
  mutate(genus = str_split(scientific_name, " ", simplify = T)[ , 1]) |>
  # Reorder columns
  relocate(genus, .before = scientific_name) |>
  # Make park names shorter
  mutate(park_name = str_remove(park_name, 'National Parks'),
         park_name = str_remove(park_name, 'National Park'),
         park_name = str_remove(park_name, 'and Preserve'))

tibble(species_clean)


parks_clean <-
  parks |>
  # Make park names shorter
  mutate(park_name = str_remove(park_name, 'National Parks'),
         park_name = str_remove(park_name, 'National Park'),
         park_name = str_remove(park_name, 'and Preserve'))

tibble(parks_clean)
```
Lastly, we'll combine the two data sets and remove the park code.

We have decided to restrict our data to the contiguous United States, for a number of reasons. Alaska and Hawaii have populations that are so different from the rest of the country that they are hardly comparable. Removing them makes our data far less complex, as it removes species that are found nowhere else in the country. It also serves the secondary purpose of making the data mapping easier. 

```{r combined dataset}
species_park <-
  species_clean |>
  # Add park info to species dataset
  left_join(y = parks_clean, by = 'park_name') |>
  # Remove park code
  select(-park_code) |>
  # Remove parks in Alaska or Hawaii for mapping purposes
  filter(!state %in% c('AK', 'HI')) 
  
tibble(species_park)
```

Here are our updated species, categories, which will be used throughout the rest of the analysis. 

```{r updated categories}

# Display updated species categories
categories_clean <-
  species_clean |>
  group_by(category) |>
  summarize(category_count = n()) |>
  select(category)

tibble(categories_clean)

```
### Data Analysis

Because our categories are not true taxonomic groupings, we need some context to the diversity within each. Lets examine orders first.

```{r overall taxanomic categories proportion}

category_prop <-
  
  species_clean |>
  group_by(category) |>
  summarize(number_cat = n()) |>
  mutate(number_tot = sum(number_cat),
         prop = number_cat/number_tot) |>
  select(category, prop)

category_prop |>
  
  ggplot(mapping = aes(
    y = prop,
    x = "",
    fill = category
  )) +

  geom_col(position = "fill",
           color = "black") +
  
  labs(x = NULL,
       y = "Proportion by Category",
       fill = "Category",
       title = "Proportion of Each Taxanomic Category")

#going to be working on this below, will probably have to do some small multiples shenanigans bc there are way too many orders
# TODO: we honestly may not need this, or at least we can wait until we get feedback from him on if we need more initial graphs/info
  species_clean |>
  group_by(category, order) |>
  summarize(number_ord = n()) |>
  mutate(number_tot = sum(number_ord),
         prop = number_ord/number_tot) |>
  select(category, order, prop) |>
  
  ggplot(mapping = aes(
    x = prop,
    y = "",
    fill = order
  )) +
    
    geom_col(position = "fill",
           color = "black") +
    labs(y = NULL,
       x = "Proportion by Order",
       fill = "Category",
       title = "Proportion of Each Order") +
    facet_wrap(facets = '~category') +
    
    # TODO: Maybe we should do number of different orders within each category rather than proportion
    # I'm trying that below
    theme(legend.position = 'none')


```
As seen, the vast majority of species observed fall into the **bird** categories. This can be explained by a few different hypotheses:

1) Because we removed seasonality, we are looking at species spotted in the park at any time of year, including species that are strictly migratory. Due to their migratory nature, the average bird species is often has a wider range of observation than that of an equivalent reptile, amphibian or mammal. 

2) Additionally, due to a variety of factors, birds are often have more comprehensive records than any other group. The MAPS program, created by the Institute for Bird Populations has gone a long way to creating distribution patterns across the United States, and the role of citizen science, on programs like eBird or iNaturalist create lists of observed species without the need for a monitoring program run by the park. 

3) Birds naturally have a high diversity, and occur across ecosystems. As such, every park is likely to have birds present, while fish, amphibians, reptiles, and mammals are far more restricted to certain habitats. 
```{r species by park}

ggplot(data = species_park |> group_by(park_name),
       mapping = aes(x = fct_infreq(park_name))) +
  
  geom_bar(fill = 'deepskyblue3') +
  
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1,
                                   size = 7),
        plot.title = element_text(size = 12,
                                  hjust = 0.5,
                                  face = 'bold'),
        axis.title = element_text(size = 10)) +
  
  labs(x = 'Park name',
       y = 'Number of species',
       fill = 'Category',
       title = 'Number of species in each park') +
  
  scale_y_continuous(expand = c(0,0, 0.05, 0)) 


````


```{r species richness order}

species_per_park <-
  species_park |>
  group_by(park_name, latitude, longitude, state, acres) |>
  summarise(num_species = n()) |>
  arrange(desc(num_species))

tibble(species_per_park)

park_order <-
  species_per_park$park_name


```


```{r taxonomic categories by park}

categories_by_park <-
  species_park |>
  group_by(park_name, category) |>
  summarise(cat_sum = n())

ggplot(data = categories_by_park,
       mapping = aes(x = factor(park_name,
                                levels = park_order))) +
  
  geom_col(mapping = aes(y = cat_sum,
                         fill = category),
           position = 'stack') +
  
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1,
                                   size = 7),
        plot.title = element_text(size = 12,
                                  hjust = 0.5,
                                  face = 'bold'),
        axis.title = element_text(size = 10)) +
  
  labs(x = 'Park name',
       y = 'Number of species',
       fill = 'Category',
       title = 'Number of species in each park by category') +
  
  scale_y_continuous(expand = c(0,0, 0.05, 0)) 



# Same plot but by proportion
ggplot(data = categories_by_park,
       mapping = aes(x = factor(park_name,
                                 levels = park_order),
                     y = cat_sum,
                     fill = category)) +
  
  geom_col(position = 'fill') +
  
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1,
                                   size = 7),
        plot.title = element_text(size = 12,
                                  hjust = 0.5,
                                  face = 'bold'),
        axis.title = element_text(size = 10)) +
  
  labs(x = 'Park name',
       y = 'Proportion of species',
       fill = 'Category',
       title = 'Proportion of species in each park by category') +
  
  scale_y_continuous(labels = scales::percent,
                     expand = c(0,0)) 
  
  

  

```
```{r map outline}
map_outline <-
  ggplot(data = map_data(map = 'state'),
         mapping = aes(x = long,
                       y = lat,
                       group = group)) +
  
  geom_polygon(fill = 'white',
               color = 'grey30') +
  
  theme_map() +
  
  coord_map(projection = "albers", 
            lat0 = 39, lat1 = 45) +
  
  scale_x_continuous(expand = c(0,0)) +
  
  scale_y_continuous(expand = c(0,0))

```


```{r species richness map}
 richness_map <-
  
  map_outline +
  
  geom_point(data = species_per_park,
           mapping = aes(x = longitude,
                        y = latitude,
                       group = park_name,
                       size = num_species,
                       color = state)) +
  
  geom_text_repel(data = species_per_park,
            mapping = aes(x = longitude,
                          y = latitude,
                          group = park_name,
                          label = park_name),
            hjust = 0.1,
            size = 1.8,
            fontface= 'bold',
            max.overlaps = 15) +
  
  scale_size(breaks = c(500, 1000, 1500, 2000, 2500, 3000)) +
  
  labs(title = 'Species richness by park',
       size = 'Number of species') +
  
  guides(color = F) +
  
  theme(legend.position = 'bottom',
        legend.direction = 'horizontal',
        legend.box.background = element_rect(color = 'black'),
        plot.title = element_text(size = 12,
                                  hjust = 0.5,
                                  face = 'bold'))
  
  
richness_map
```


```{r park size map}
richness_map <-
  
  map_outline +
  
  geom_point(data = species_per_park,
           mapping = aes(x = longitude,
                        y = latitude,
                       group = park_name,
                       size = acres,
                       color = state)) +
  
  geom_text_repel(data = species_per_park,
            mapping = aes(x = longitude,
                          y = latitude,
                          group = park_name,
                          label = park_name),
            hjust = 0.1,
            size = 1.8,
            fontface= 'bold',
            max.overlaps = 15) +
  
  # TODO: Should figure out better breaks maybe?
  scale_size(breaks = seq(from = 100000, to = 2500000, by = 200000)) +
  
  labs(title = 'Park size',
       size = 'Acreage') +
  
  guides(color = F) +
  
  theme(legend.position = 'bottom',
        legend.direction = 'horizontal',
        legend.box.background = element_rect(color = 'black'),
        plot.title = element_text(size = 12,
                                  hjust = 0.5,
                                  face = 'bold'))
  
  
richness_map
```

```{r scatter plot of size vs richness}
# very much in progress duh
ggplot(data = species_per_park,
       mapping = aes(x = acres,
                     y = num_species)) +
  
  geom_point()
```

```{r size richness correlation}
species_per_park |>
  mutate(
    number = "num_species"
  ) -> 
  species_corr 
  


ggcorr(
  data = species_corr,
  low = "red3",
  mid = "white",
  high = "blue3",
  label = T,
  label_round = 2
  
)
```

>>>>>>> 3917d00f22373b55a84495ac7d875e18065b79e2
